\vspace{-1cm} \bf Approximate DP and Regret Analysis \vspace{-1.3cm}

### 2018-07-12 Draft.tex
Relaxed Bellman equations

\section{Probing II: Beyond Bernoulli and formulation as a MCDP}

@@

The state space for our problem can be described as $\St=\crl{(b_h,b_p),(b_h,b_p,\diamond):b_h,b_p\in\N,\diamond\in \crl{\texttt{a,p,r}}}$, see \cref{fig:probing_states}.
The component $\diamond$ indicates if we are in the accept, probe or reject stage.
There is never a $\diamond$ at times $2t$ and, at $2t-1$, feasible controls are different for distinct $\diamond$.

@@

@@

Let $\calX\defeq\Rp^{4n}$ be the space for the decision variables.
Note that, by definition, $Z_j(2t-1)=Z_j(2t)-\In{\xi^{2t}=j}$.
In \cref{eq:probing_even,eq:probing_odd} we present two similar functions, observe that they differ by the last two constraints only.

\begin{equation}\label{eq:probing_even}
\begin{array}{rrll}
\varphi(2t,b,Z,\xi^{2t}) \; \max & \multicolumn{3}{l}{\sum_{j,k}r_{jk}(x_{(j,k)a}+q_{jk}x_{ja})} \\
\text{s.t.}& \sum_j x_{jp} & \leq b_p \\
& \sum_{j,k}x_{(j,k)a}+\sum_jx_{ja} &\leq b_h \\
&  x_{ja} +x_{jp} & \leq Z_j(t)  & j\in [n] \\
& x_{(j,k)a} &\leq q_{jk}x_{jp} &  j \in [n],k\in [2].
\end{array}
\end{equation}
\begin{equation}\label{eq:probing_odd}
\begin{array}{rrll}
\varphi(2t-1,b,Z,\xi^{2t-1}) \; \max & \multicolumn{3}{l}{\sum_{j,k}r_{jk}(x_{(j,k)a}+q_{jk}x_{ja})} \\
\text{s.t.}& \sum_j x_{jp} & \leq b_p \\
& \sum_{j,k}x_{(j,k)a}+\sum_jx_{ja} &\leq b_h \\
&  x_{ja} + x_{jp} & \leq Z_j(t)-\In{\xi^{2t}=j}  & j\in [n] \\
& x_{(j,k)a} &\leq q_{jk}x_{jp} + \In{\xi^{2t-1}=(j,k)}&  j \in [n],k\in [2].
\end{array}
\end{equation}

An argument as in \cref{prop:probing_relaxation} shows that $\voff(t,b)\leq \varphi(t,b,Z,\xit)$.
We can pretend, thus, that \off agrees to receive rewards according to $\varphi$.
To make this precise, we need a map between two worlds, since \onl decides on actions \texttt{a,p,r} and \off on points in $\calX$.

\begin{lemma}
Let $\bar X\in\calX$ be a maximizer of $\varphi(2t,b,Z,\xi^{2t})$.
Say $\xi^{2t}=j$, then, at time $2t$, \off is satisfied with the following actions:
\begin{enumerate}
\item Accept if $\bar X_{ja}\geq 1$
\item Probe if $\bar X_{jp}\geq 1$
\item Reject if $\bar X_{ja}+\bar X_{jp}\leq Z_j(t)-1$
\end{enumerate}
At time $2t-1$, \off is satisfied accepting type $(j,k)$ in a probing stage if $\bar X_{jp}\geq 1$ and $\bar X_{(j,k)a}\geq 1$.
\end{lemma}
\begin{proof}
\off is satisfied solving the following problem, where the actions correspond, respectively, to accept, probe and reject
\[
\max\crl{\bar r_j+\E_{\xi^{2t-2}}[\varphi(2t-2,b-e_h,Z,\xi^{2t-2})], \E_{\xi^{2t-2}}[\varphi(2t-1,b-e_p,Z,\xi^{2t-1})], \E_{\xi^{2t-2}}[\varphi(2t-2,b,Z,\xi^{2t-2})]}.
\]
For the accept and reject options the expectation is redundant, since all the necessary information to solve the problem is given in $Z$.
This is not true for the probe option, since \off is unaware of $\xi^{2t-2}$.
By Jensen's inequality, $\E_{\xi^{2t-2}}[\varphi(2t-1,b-e_p,Z,\xi^{2t-1})]\leq \varphi(2t-1,b-e_p,Z,\E[\xi^{2t-1}])$, where we are abusing notation to denote that in \cref{eq:probing_odd} we are replacing $\In{\xi^{2t-1}=(j,k)}$ by $q_{jk}$.
It follows now that \off is also satisfied solving the problem 
\[
\max\crl{\bar r_j+\varphi(2t-2,b-e_h,Z,\xi^{2t-2}), \varphi(2t-1,b-e_p,Z,\E[\xi^{2t-1}])], \varphi(2t-2,b,Z,\xi^{2t-2})}.
\]
Finally, we can relate each of these options with the optimum of $\varphi(2t,b,Z,\xi^{2t})$ in the cases (1), (2) and (3).
The argument at odd times $2t-1$ is similar. 
\end{proof}

@@

Consider a DP with state space $\St$ and a sequence of inputs $\xit\in\Xi$ for $t\in[\T]$, where $t$ is the \emph{time to-go}.
We assume $\xit$ evolves as a Markov chain on state space $\Xi$, with known transition matrix and initial distribution.

For every state $s\in\St$, arrival $\xi$ and control $u\in\U$, we  collect a reward $\Re(s,\xi,u)$, which could be $-\infty$, and transition to the state $\Tr(s,\xi,u)$.
Observe that the reward function is deterministic, but the input $\xi$ is random, thus making the reward itself at each time a random variable.
We stress that, at each time $t$, the decision-maker first observes $\xit$ and based on that makes a decision.
\anote{
The only difference is that in MCDP the reward depends on a pair of states (current and transition), but I think that would be redundant.
}
Any time dependence will be embedded in the process $\xit$, e.g., if the randomness represents unknown demand, the state of the Markov chain could be of the form $\J=\crl{(t,x):t\in[\T],x\in\Rp}$, where the first component gives us the time and the second the demand for the period.
This formulation is known as a Markov Chance-Decision Process \cite[chap 13]{online_book}.

Given this formulation, the value function at any given time can be expressed as
\begin{equation}
v(t,s) = \E_{\xit,\ldots,\xi^1}\brk*{\max_{u^t,\ldots,u^1}\sum_{\tau=1}^t\Re(S^\tau,\xi^\tau,u^\tau) \text{ s.t. }  S^t=s \text{ and } S^{\tau-1}=\Tr(S^\tau,\xi^\tau,u^\tau) \forall \tau<t }.
\end{equation}

@@

\paragraph{Assumption: Finite Dimension}
We identify $\St\subseteq \R^\ds$ and $\U\subseteq\R^\du$.

@@

Similarly, there is a mapping $I:\St\to [m]$ and we denote $\It=I(S^t)$.
We assume there is a convex function $f:\R^\ds\times\R^\du\to\R^d$, called \emph{feasibility function}, and discrete sets $\U_{ij}$ such that
\[
\U(s,\xi) = \crl{u\in \U_{I(s),J(\xi)}: f(s,u)\geq 0}.
\]

>>>>

\section{Probing with Multiple Values and Formulation as a MCDP}

@@

The state space for our problem can be described as $\S=\crl{(b_h,b_p),(b_h,b_p,\diamond):b_h,b_p\in\N,\diamond\in \crl{\texttt{a,p,r}}}$, see \cref{fig:probing_states}.
The component $\diamond$ indicates if we are in the accept, probe or reject stage.
There is never a $\diamond$ at times $2t$ and, at $2t-1$, feasible controls are different for distinct $\diamond$.

@@

\section{Relaxed Bellman Equations}

Consider a DP with state space $\S$ and a sequence of inputs $\xit\in\Xi$ for $t\in[\T]$, where $t$ is the \emph{time to-go}.
We assume $\xit$ evolves as a Markov chain on state space $\Xi$, with known transition matrix and initial distribution.

For every state $s\in\S$, arrival $\xi$ and control $u\in\U$, we  collect a reward $\Re(s,\xi,u)$, which could be $-\infty$, and transition to the state $\Tr(s,\xi,u)$.
Observe that the reward function is deterministic, but the input $\xi$ is random, thus making the reward itself at each time a random variable.
We stress that, at each time $t$, the decision-maker first observes $\xit$ and based on that makes a decision.
\anote{
The only difference between this formulation and MCDP is that in MCDP the reward depends on a pair of states (current and transition), but I think that would be redundant.
}
Any time dependence will be embedded in the process $\xit$, e.g., if the randomness represents unknown demand, the state of the Markov chain could be of the form $\J=\crl{(t,x):t\in[\T],x\in\Rp}$, where the first component gives us the time and the second the demand for the period.
This formulation is known as a Markov Chance-Decision Process \cite[chap 13]{online_book}.

Given this formulation, if at time $t$ the current state is $s\in\S$, we obtain the Bellman \cref{eq:bellman_mcdp}
\begin{equation}\label{eq:bellman_mcdp}
v(t,s,\xit) = \max_{u\in\U}\crl{\Re(s,\xit,u) +\E_{\xi^{t-1}}[v(t-1,\Tr(s,\xit,u),\xi^{t-1})|\xit]},
\end{equation}
with boundary condition $v(0,\cdot,\cdot)=0$ and $\xi^0$ set to an arbitrary value.
This is a well-known fact; contingent on the actions being optimal, the total to-go value can be decomposed as instant reward plus future value.
We will now state a relaxed version of \cref{eq:bellman_mcdp}, where the main idea is to replace $v$, which is difficult to compute, for an efficiently computable upper bound $\tilde v$.
If $\tilde v$ satisfies a probabilistic version of a Bellman Equation, then the process of iteratively solving the relaxed problem can be justified.

\begin{definition}[Relaxed Bellman Equations]
The function $\varphi:[T]\times\S\times\Xi\to \R$ satisfies the relaxed Bellman equations w.r.t.\ $v$ if:
\begin{enumerate}
\item Initial condition: $\E[v(\T,S^\T,\xi^\T)] \leq \E[\varphi(\T,S^\T,\xi^\T)]$
\item Probabilistic inequality: $\forall s\in\S,t\in[T]$, there is a set $\calB(t,s)\subseteq\Omega$ such that
\begin{equation}\label{eq:relaxed_bellman}
\varphi(t,s,\xit) \leq \max_{u\in\U}\crl{R(s,\xit,u)+\E_{\xi^{t-1}}[\varphi(t-1,\Tr(s,\xit,u),\xi^{t-1})|\xit]} \quad \forall \omega\in\calB(t,s).
\end{equation}
\end{enumerate}
\end{definition}

\begin{example}
Consider a knapsack problem with two types of stochastic arrivals.
Type $j$ has reward $r_j$, arrives at each period with probability $p_j$ and uses space $w_j$ in the knapsack.
Assume all the arrivals are revealed.
We will show that the following satisfies the relaxed Bellman equations:
\[
\varphi(t,b,\xit) = \max\crl*{\sum_{j}x_jr_j:\sum_{j}w_jx_j\leq b, \quad 0\leq x_j \leq Z_j(t) \quad \forall j\in[2]}.
\]
The initial condition is trivial, since we are relaxing the integrality constraints.
We are left to show the probabilistic inequality.
Let $X[\omega]$ be a solution to the LP and let $i=\xit$.
For the remainder of the example, we will assume $X_i\in(0,1)$.
If $Z_i(t)\geq 2$,  then an exchange argument shows
\[
\varphi(t,b,\xit) = \max_{u=\texttt{r,a}}\crl{r(b,\xit,u)+\varphi(t-1,b\ominus u,\xi^{t-1})}
= \varphi(t-1,b,\xi^{t-1}),
\]
where $b\ominus u$ denotes the natural reduction of budget of an action.
We conclude that, if $Z_i(t)\geq 2$ and $X_i<1$, the probabilistic inequality is satisfied a.s.
We are left with the case $Z_i(t)=1$ (the case $Z_i(t)=0$ is impossible when $X_i\in (0,1)$).
In this case, 
\[
\varphi(t,b,\xit) > \max_{u=\texttt{r,a}}\crl{r(b,\xit,u)+\varphi(t-1,b\ominus u,\xi^{t-1})}.
\]
We conclude that the probabilistic inequality is satisfied in $\calB(t,b)=\crl{\omega\in\Omega:Z_i(t)\geq 2}$.
Observe that the probability of this set is exponentially close to one.
Indeed, $\Pr[\calB(t,b)]=1-tp_i(1-p_i)^{t-1}$.
\end{example}

\begin{proposition}\label[proposition]{prop:relaxed_bellman}
Assume $\varphi$ satisfies the relaxed Bellman equations w.r.t.\ $v$.
Denote $\St$ the state of the policy that takes action $\Ut=\Ut(\xit,\St)$, where $\Ut$ is a maximizer in \cref{eq:relaxed_bellman}.
Then,
\[
\E[v(\T,S^\T,\xi^\T)] \leq \E\brk*{ \sum_{t=1}^\T (\Re(\St,\xit,\Ut)+\rmax\Ins{B(t,\St)})}. 
\]
\end{proposition}
\begin{proof}
By the initial condition property, $\E[v(\T,S^\T,\xi^\T)] \leq \E[\varphi(\T,S^\T,\xi^\T)]$.
It is now enough to argue the following equation and conclude by a simple induction step.
\[
\varphi(t,\St,\xit)\leq \E_{\xi^{t-1}}[\Re(\St,\xit,\Ut)+\varphi(t-1,S^{t-1},\xi^{t-1}) +  \rmax\Ins{B(t,\St)}|\xit].
\]
By definition of $\Ut$, $\varphi(t,\St,\xit)\leq \Re(\St,\xit,\Ut)+\E_{\xi^{t-1}}[\varphi(t-1,S^{t-1},\xi^{t-1})] +  \rmax\Ins{B(t,\St)}$.
Since $\Re(\St,\xit,\Ut)$ and $\Ins{B(t,\St)}$ are measurable with respect to $\xit$, they can be inside the expectation, showing that the desired equation holds.
\end{proof}

\anote{The $\rmax$ term here corresponds to the loss in $\varphi$.
All the relaxations we use have the same $\rmax$ as we would expect.}

\section{Relaxation for Multiple Probing}

@@

Let $\calX\defeq\Rp^{4n}$ be the space for the decision variables.
In \cref{eq:probing_lp} we present a LP which will be the building block for our relaxation.
Note that, by definition, if $t$ is odd, $Z_j(t)=Z_j(t+1)-\In{\xi^{t+1}=j}$.

\begin{equation}\label{eq:probing_lp}
\begin{array}{rrll}
P[t,Z,b] \;= \max & \multicolumn{3}{l}{\sum_{j,k}r_{jk}(x_{(j,k)a}+q_{jk}x_{ja})} \\
\text{s.t.}& \sum_{j,k}x_{(j,k)a}+\sum_jx_{ja} &\leq b_h  \\
&  \sum_j x_{jp} & \leq b_p   \\
&  x_{ja} +x_{jp} & \leq Z_j(t)  & j\in [n] \\
& x_{(j,k)a} &\leq q_{jk}x_{jp} &  j \in [n],k\in [2]\\ 
& x&\in\calX.
\end{array}
\end{equation}

The LP in \cref{eq:probing_lp} can be interpreted as follows...\todonote

We are ready to parse the relaxation in \cref{eq:probing_relax}.
Recall that a state is of the form $s=(b_h,b_p,\diamond)$ with $\diamond\in \crl{\texttt{a},\texttt{p},\texttt{r},\varnothing}$, where $\diamond=\varnothing$ for even times.
\begin{equation}\label{eq:probing_relax}
\varphi(t,s,\xit) = \left\{ 
\begin{array}{ll}
v(P[t,Z,b]) & \diamond=\varnothing \\ 
v(P[t-1,Z,b]) & \diamond = \texttt{r} \\ 
r_{\xit}+v(P[t-1,Z,b]) & \diamond = \texttt{a} \\ 
\max\crl{r_{\xit}+v(P[t-1,Z,b-e_h]),v(P[t-1,Z,b])} & \diamond = \texttt{p}
\end{array} 
\right.
\end{equation}

\begin{lemma}
Let $\bar X\in\calX$ be a maximizer of $(P[t,Z,b])$ for $t$ even.
Say $\xit=j$, then, at time $t$, \off is satisfied with the following actions:
\begin{enumerate}
\item Accept if $\bar X_{ja}\geq 1$
\item Reject if $\bar X_{ja}+\bar X_{jp}\leq Z_j(t)-1$
\item Probe if $\bar X_{jp}\geq 1$ and then accept $(j,k)$ if $\bar X_{(j,k)a}\geq 1$
\end{enumerate}
\end{lemma}
\begin{proof}
We will show that $\varphi$ satisfies the relaxed Bellman equations, then conclude in virtue of \cref{prop:relaxed_bellman}.
The initial condition follows from the same argument as in \cref{prop:probing_relaxation}.
We are left to show the the probabilistic inequality
\[
\varphi(t,s,\xit) \leq \max_{u\in\U}\crl{R(s,\xit,u)+\E_{\xi^{t-1}}[\varphi(t-1,\Tr(s,\xit,u),\xi^{t-1})|\xit]} \quad \forall \omega\in\calB(t,s).
\]
The set $\calB(t,s)$ is defined as the $\omega\in\Omega$ such that at least one of the conditions (1),(2) or (3) are satisfied.
Observe that, since $t$ is even, the instant reward $R(s,\xit,u)$ is zero.
In cases (1) and (2) is easy to verify the probabilistic inequality.
For case (3) we apply Jensen's Inequality:
\begin{align*}
\varphi(t,s,\xit) &= v(P[t,Z,b]) \\
&= \max\crl{\E[r_{\xi^{t-1}}|\xit]+v(P[t-1,Z,b-e_p-e_h]),v(P[t-1,Z,b-e_p])} \\
&\leq \E_{\xi^{t-1}}[\max\crl{r_{\xit}+v(P[t-1,Z,b-e_h-e_p]),v(P[t-1,Z,b-e_p])}].
\end{align*}
The last term equals $R(s,\xit,u)+\E_{\xi^{t-1}}[\varphi(t-1,s\ominus\texttt{p},\xi^{t-1})|\xit]$ and the proof is complete.
\end{proof}

@@

Iterating the Bellman \cref{eq:bellman_mcdp}, we arrive at the equivalent expression 
\begin{equation}
v(t,s,\xit) = \E_{\xi^{t-1},\ldots,\xi^1}\brk*{\max_{u^t,\ldots,u^1}\sum_{\tau=1}^t\Re(S^\tau,\xi^\tau,u^\tau) \text{ s.t. }  S^t=s \text{ and } S^{\tau-1}=\Tr(S^\tau,\xi^\tau,u^\tau) \forall \tau<t }.
\end{equation}

@@

\paragraph{Assumption: Finite Dimension}
We identify $\S\subseteq \R^\ds$ and $\U\subseteq\R^\du$.

@@

Similarly, there is a mapping $I:\S\to [m]$ and we denote $\It=I(S^t)$.
We assume there is a convex function $f:\R^\ds\times\R^\du\to\R^d$, called \emph{feasibility function}, and discrete sets $\U_{ij}$ such that
\[
\U(s,\xi) = \crl{u\in \U_{I(s),J(\xi)}: f(s,u)\geq 0}.
\]
